---
title: "Differential Expression Analysis"
author: "Rob Schmidt"
output: html_notebook
---

```{r setup}
library(tidyverse)
library(DESeq2)
```


```{r exploration, message=FALSE}
sampleData<-read_tsv("../SampleTable.txt")%>%
  mutate(Time=as.factor(Time), Antibiotic=as.factor(Antibiotic),
         group=as.factor(paste(Antibiotic, Time, sep="_")))

sampleData$Antibiotic<-relevel(sampleData$Antibiotic, ref="NDC")
sampleData$group<-relevel(sampleData$group, ref="NDC_30")


files<-list.files("../HTSeqOut/Gene_level/", full.names = T)

ctMat<-map(files, read_tsv, col_names=c("GeneID", "Expression"))%>%
  purrr::reduce(full_join, by="GeneID")%>%
  filter(GeneID != "__no_feature",  GeneID != "__ambiguous", 
         GeneID != "__too_low_aQual",  GeneID != "__not_aligned", 
         GeneID != "__alignment_not_unique")%>%
  column_to_rownames("GeneID")%>%
  as.matrix()
colnames(ctMat)<-sampleData$Run

dds<-DESeqDataSetFromMatrix(ctMat, sampleData, design=~Time*Antibiotic)



normCts<-counts(dds, normalized=F)%>%
  rlog()

normCts_long<-normCts%>%
  as.data.frame()%>%
  rownames_to_column("GeneID")%>%
  pivot_longer(-GeneID, names_to="Sample", values_to="NormCounts")
```

```{r Figure1, caption="Figure 1: Distribution of rlog transformed counts by sample"}
#Count distributions
ggplot(normCts_long, aes(x=Sample, y=log10(NormCounts)))+
  geom_boxplot()
```
```{r}
ggplot(normCts_long, aes(x=log10(NormCounts+1e-7), fill=Sample))+
  geom_density(alpha=0.1)+
  theme(legend.position = "none")
```

```{r}
#Count distributions look ok.

#PCA

pca<-prcomp(t(normCts))

cumVar<-pca$sdev^2/sum(pca$sdev^2)*100

pcaPlotDat<-pca$x%>%
  as.data.frame()%>%
  rownames_to_column("Run")%>%
  inner_join(sampleData, by="Run")

ggplot(pcaPlotDat, aes(x=PC1, y=PC2, shape=Time, color=Antibiotic))+
  geom_point()+
  labs(x=paste("PC1", paste0("[", paste0(as.character(round(cumVar[1], 2)), "%"),"]"), sep=" "), y=paste("PC2", paste0("[", paste0(as.character(round(cumVar[2], 2)), "%"),"]"), sep=" "))+
  theme_classic() #Time looks to be dominate difference. Antibiotic does differ, but not as much as time. Rockstar antibiotics seem to be TOB, CIP, MER, COL as these seperate fairly well.
```




```{r ModelFitting}

#Check simpler model
dds2<-DESeq(dds, test="LRT", reduced=~Antibiotic+Time)

res<-results(dds2, tidy=T)%>%
  filter(padj < 0.05)

#1119 genes out of 3764 genes have a greater increase in the log liklihood than would be expected if the interaction term was 0. That's about 30% of the observed genes that the full model may fit better to. Therefore I will continue with the full model. 

```

```{r resultsSummary}
dds2<-DESeq(dds)
```

```{r}
plotDispEsts(dds2) #Fit is good with high mean counts. > 0.5?
```


##Cip Results
```{r}
resultsNames(dds2)
#genes changing with CIP adjusted for NDC should be (CIP90-NDC90)-(CIP30-NDC30)+(NDC90-NDC30) = CIP90 - CIP30
#Ought to be just the interaction term for CIP plus the time term.

CIPRes<-results(dds2, contrast=c(0,1,0,0,0,0,0,0,1,0,0,0,0,0), tidy=T)

CIPSig<-CIPRes%>%filter(padj < 0.05, abs(log2FoldChange) >= 1.5)

write_tsv(CIPRes, "../DESeq2Out/ByAntibioticResults/FullResults/CIP_FullResults.txt")
write_tsv(CIPSig, "../DESeq2Out/ByAntibioticResults/SignificantResults/CIP_SigResults.txt")
```
```{r}
#pvals
ggplot(CIPRes, aes(x=pvalue))+geom_histogram()+
  scale_y_discrete(expand=c(0,0))+
  theme_classic()
```
```{r}
#volcano plot
CIPResVPlot<-CIPRes%>%
  mutate(Significance=ifelse(padj<0.05 & abs(log2FoldChange) >= 1.5, "Significant", "Not Significant"), alpha=ifelse(Significance == "Not Significant", 0.1, 1))%>%
  filter(!is.na(padj))
ggplot(CIPResVPlot, aes(x=log2FoldChange, y=-log10(padj), color=Significance, alpha=alpha))+
  geom_point()+
  scale_color_manual(values=c("black", "red"))+
  labs(title="CIP", x="Log2 Fold Change", y="-log10(adjusted p-value)")+
  theme_classic()+
  theme(legend.position = "none")
```

In total there are 368 genes that significantly differ between incubation with CIP for 90 mins incubation vs CIP for 30 min. Significane is defined as having a BH adjusted p value less than 0.05 and the absolute value of the log2 fold change is greater than 1.5.

##COL Analysis
```{r, COLAnalysis}
COLRes<-results(dds2, contrast=c(0,1,0,0,0,0,0,0,0,1,0,0,0,0), tidy=T)

COLSig<-COLRes%>%filter(padj < 0.05, abs(log2FoldChange) >= 1.5)

write_tsv(COLRes, "../DESeq2Out/ByAntibioticResults/FullResults/COL_FullResults.txt")
write_tsv(COLSig, "../DESeq2Out/ByAntibioticResults/SignificantResults/COL_SigResults.txt")
```

```{r}
#pvals
ggplot(COLRes, aes(x=pvalue))+geom_histogram()+
  scale_y_discrete(expand=c(0,0))+
  theme_classic()
```

```{r}
#volcano plot
COLResVPlot<-COLRes%>%
  mutate(Significance=ifelse(padj<0.05 & abs(log2FoldChange) >= 1.5, "Significant", "Not Significant"), alpha=ifelse(Significance == "Not Significant", 0.1, 1))%>%
  filter(!is.na(padj))
ggplot(COLResVPlot, aes(x=log2FoldChange, y=-log10(padj), color=Significance, alpha=alpha))+
  geom_point()+
  scale_color_manual(values=c("black", "red"))+
  labs(title="COL", x="Log2 Fold Change", y="-log10(adjusted p-value)")+
  theme_classic()+
  theme(legend.position = "none")
```

There are a total of 40 genes that are DE with colsitin.

##IMI Analysis

```{r}
IMIRes<-results(dds2, contrast=c(0,1,0,0,0,0,0,0,0,0,1,0,0,0), tidy=T)

IMISig<-IMIRes%>%filter(padj < 0.05, abs(log2FoldChange) >= 1.5)

write_tsv(IMIRes, "../DESeq2Out/ByAntibioticResults/FullResults/IMI_FullResults.txt")
write_tsv(IMISig, "../DESeq2Out/ByAntibioticResults/SignificantResults/IMI_SigResults.txt")
```

```{r}
#pvals
ggplot(IMIRes, aes(x=pvalue))+geom_histogram()+
  scale_y_discrete(expand=c(0,0))+
  theme_classic()
```

```{r}
#volcano plot
IMIResVPlot<-IMIRes%>%
  mutate(Significance=ifelse(padj<0.05 & abs(log2FoldChange) >= 1.5, "Significant", "Not Significant"), alpha=ifelse(Significance == "Not Significant", 0.1, 1))%>%
  filter(!is.na(padj))
ggplot(IMIResVPlot, aes(x=log2FoldChange, y=-log10(padj), color=Significance, alpha=alpha))+
  geom_point()+
  scale_color_manual(values=c("black", "red"))+
  labs(title="IMI", x="Log2 Fold Change", y="-log10(adjusted p-value)")+
  theme_classic()+
  theme(legend.position = "none")
```

##MER Analysis

```{r}
MERRes<-results(dds2, contrast=c(0,1,0,0,0,0,0,0,0,0,0,1,0,0), tidy=T)

MERSig<-MERRes%>%filter(padj < 0.05, abs(log2FoldChange) >= 1.5)

write_tsv(MERRes, "../DESeq2Out/ByAntibioticResults/FullResults/MER_FullResults.txt")
write_tsv(MERSig, "../DESeq2Out/ByAntibioticResults/SignificantResults/MER_SigResults.txt")
```

```{r}
#pvals
ggplot(MERRes, aes(x=pvalue))+geom_histogram()+
  scale_y_discrete(expand=c(0,0))+
  theme_classic()
```

```{r}
#volcano plot
MERResVPlot<-MERRes%>%
  mutate(Significance=ifelse(padj<0.05 & abs(log2FoldChange) >= 1.5, "Significant", "Not Significant"), alpha=ifelse(Significance == "Not Significant", 0.1, 1))%>%
  filter(!is.na(padj))
ggplot(MERResVPlot, aes(x=log2FoldChange, y=-log10(padj), color=Significance, alpha=alpha))+
  geom_point()+
  scale_color_manual(values=c("black", "red"))+
  labs(title="MER", x="Log2 Fold Change", y="-log10(adjusted p-value)")+
  theme_classic()+
  theme(legend.position = "none")
```

##MOX Analysis

```{r}
MOXRes<-results(dds2, contrast=c(0,1,0,0,0,0,0,0,0,0,0,0,1,0), tidy=T)

MOXSig<-MOXRes%>%filter(padj < 0.05, abs(log2FoldChange) >= 1.5)

write_tsv(MOXRes, "../DESeq2Out/ByAntibioticResults/FullResults/MOX_FullResults.txt")
write_tsv(MOXSig, "../DESeq2Out/ByAntibioticResults/SignificantResults/MOX_SigResults.txt")
```

```{r}
#pvals
ggplot(MOXRes, aes(x=pvalue))+geom_histogram()+
  scale_y_discrete(expand=c(0,0))+
  theme_classic()
```

```{r}
#volcano plot
MOXResVPlot<-MOXRes%>%
  mutate(Significance=ifelse(padj<0.05 & abs(log2FoldChange) >= 1.5, "Significant", "Not Significant"), alpha=ifelse(Significance == "Not Significant", 0.1, 1))%>%
  filter(!is.na(padj))
ggplot(MOXResVPlot, aes(x=log2FoldChange, y=-log10(padj), color=Significance, alpha=alpha))+
  geom_point()+
  scale_color_manual(values=c("black", "red"))+
  labs(title="MOX", x="Log2 Fold Change", y="-log10(adjusted p-value)")+
  theme_classic()+
  theme(legend.position = "none")
```

##TOB Analysis

```{r}
TOBRes<-results(dds2, contrast=c(0,1,0,0,0,0,0,0,0,0,0,0,0,1), tidy=T)

TOBSig<-TOBRes%>%filter(padj < 0.05, abs(log2FoldChange) >= 1.5)

write_tsv(TOBRes, "../DESeq2Out/ByAntibioticResults/FullResults/MOX_FullResults.txt")
write_tsv(TOBSig, "../DESeq2Out/ByAntibioticResults/SignificantResults/MOX_SigResults.txt")
```

```{r}
#pvals
ggplot(TOBRes, aes(x=pvalue))+geom_histogram()+
  scale_y_discrete(expand=c(0,0))+
  theme_classic()
```

```{r}
#volcano plot
TOBResVPlot<-TOBRes%>%
  mutate(Significance=ifelse(padj<0.05 & abs(log2FoldChange) >= 1.5, "Significant", "Not Significant"), alpha=ifelse(Significance == "Not Significant", 0.1, 1))%>%
  filter(!is.na(padj))
ggplot(TOBResVPlot, aes(x=log2FoldChange, y=-log10(padj), color=Significance, alpha=alpha))+
  geom_point()+
  scale_color_manual(values=c("black", "red"))+
  labs(title="TOB", x="Log2 Fold Change", y="-log10(adjusted p-value)")+
  theme_classic()+
  theme(legend.position = "none")
```
```{r DESummaryTable}

```

```{r Annotation}
gffID_to_geneNames<-read_tsv("../gffID_to_GeneID.txt")

names<-c("CIP", "COL", "IMI", "MER", "MOX", "TOB")
DERes<-list(CIPSig, COLSig, IMISig, MERSig, MOXSig, TOBSig)%>%
  map(inner_join, gffID_to_geneNames, by=c("row"="GeneID"))

names(DERes)<-names
```